---
title: "Collaborative Data Science Project"
format: html
author: 
  - Juan Pablo Lozano Peña
  - Megan Do
  - Carly Dempsey
editor_options: 
  chunk_output_type: console
---

```{r}
#| output: false

knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)

# load in required packages
library(tidyverse)
library(vegan)
library(BiodiversityR)
library(modelr)
library(tidybayes)
library(ggplot2)
library(tibble)
```

## Research Question

##### *How does benthic cover vary across three reef zones of the Belize Barrier Reef?*

## Visualization of Data

#### 1) Metadata

This dataset contains the metadata for quadrat photos taken every 5m along 50m transects. Six transects were conducted at each of the three study sites in Belize: lagoon (whale shoal), fore reef, and back reef. The dataset contains the **image name**, **site of collection**, **date of collection**, **transect number**, and **quadrat number**.

```{r}
metadata <- read.csv("data/Belize_CoralNet_metadata.csv")
head(metadata)
```

#### 2) Percent Cover Dataset

This dataset consists of percent cover data for each quadrat photo. Percent cover was calculated using CoralNet, a software that automates the analysis of coral reef benthic images. Essentially, CoralNet generates 25 random points over each image, identifies/labels the points, and then calculates percent cover from that data. The dataset contains the **image ID**, **image name** (same as above), **annotation status** (whether the image has been labeled), **number of collection points**, and various columns corresponding to **different benthic categories** such as hard coral species, soft coral species, macroalgal species, etc. See the [Glossary](#glossary) for a key to each category.

```{r}
per.cover.data <- read.csv("data/Belize_CoralNet_percent.covers.csv")
head(per.cover.data)
```

## Data Wrangling

Combine the metadata with the percent cover data.

```{r}
benthic.data <- metadata %>%
  left_join(per.cover.data)
```

Now, we have to calculate the average percent cover for each category at each transect. To do this, we first must `gather()` our data so that we have a column for the categories and a column for the percent cover of each category. At this time, we will also add a new column with the names of the reef zones: lagoon, forereef, backreef.

```{r}
benthic.data <- benthic.data %>%
  gather(key = category, value = percent_cover, 9:69) %>% # gather all category columns into one column, "category"
  mutate(Habitat = case_when(Site == "BZE_CBC_BACKREEF" ~ "Backreef",
                            Site == "BZE_CBC_FOREREEF" ~ "Forereef",
                            Site == "BZE_whaleShoal" ~ "Lagoon")) # add a column with habitat types
```

Next, we can group by Habitat, Transect, and Category to calculate the average percent cover for each category at each transect across the ten quadrats.

```{r}
category.cover <- benthic.data %>%
  group_by(Habitat, Transect, category) %>%
  summarize(transect_cover = sum(percent_cover)/1000) # divide by 1000 because there are 10 quadrats in a single transect: 100% * 10
head(category.cover)
```

This dataset now shows the proportion of each category found in each transect. However, we also want to classify the categories into broad functional groups for downstream analyses.

```{r}
# classify functional groups
func.group.cover <- category.cover %>%
  mutate(func_group = case_match(
    category,
    
    # macroalgae
    c("Brown_UFM", "Green_UFM", "Red_UFM", "Dictyota", "Halimeda", "Ulva", "Lobophora", "Sargassum", "Mac_matrix",
      "Brown_Enc", "Red_Enc") ~ "Macroalgae",
    
    # corals
    c("AcrCerv", "AcrPalm", "AgAga", "AgTen", "AgUnd", "BL_OrbFav", "BORB", "CoNat", "DiplLab", "FaviaFrag", "MontaCav", "OrbAnn", "OrbFav", "OrbFrank", "PorAstr", "Por_branch", "PorPor", "PseuCliv", "PseuStrig", "Siderastr", "SidSid") ~ "Coral",
    
    # gorgonians
    c("Gorg_Fan", "Gorg_Plume", "Gorg_Whip", "GORG-BRI") ~ "Gorgonian",
    
    # EAM
    c("EAM_HR", "EAM_LR", "EAM_RBL") ~ "EAM",
    
    #Cyanobacteria
    c("Cyano_HR", 
      "Cyano_LR", "Cyano_RBL", "Cyano_Sand") ~ "Cyanobacteria",
    
    # crustose coralline algae (CCA)
    c("CCA_LR", "CCA_HR", "CCA_RBL") ~ "CCA",
    
    # sponges
    c("Spong_Crus", "Spong_Erec", "Spong_Glo") ~ "Sponge",
    
    # seagrass
    c("SyrFil", "ThalTes") ~ "Seagrass",
    
    # all other categories
    .default = "Other"))


# calculate percent cover for each functional group in each transect
func.group.cover <- func.group.cover %>%
  select(-category) %>%
  group_by(Habitat, Transect, func_group) %>%
  summarise(transect_cover = sum(transect_cover)) # use `sum()` because different categories were grouped
head(func.group.cover)
```

After calculating the percent cover of each functional group for each transect, we need to format the data into "wide" format using `spread()` for downstream analyses. Thus, each row is an observation and each column is a unique variable. 

```{r}
benthic.wide <- func.group.cover %>% 
  spread(key = func_group, value = transect_cover, fill = 0) # set missing values as "0"
head(benthic.wide)
```

## Principal Coordinate Analysis
Principal Coordinate Analysis (PCoA) is a technique used to visualize multidimensional data. Since we have benthic cover data (functional group abundances from 0-1), we can use the Manhattan distance measure to calculate pairwise distances between each transect in our dataset. This will, ultimately, form a distance matrix that we can then analyze using PCoA to determine how similar or different the benthic cover of each transect is to one another.  
*Note: The Bray-Curtis distance measure could also be used for abundance data, but here we chose the Manhattan distance to analyze the absolute differences in percent cover, rather than relative metrics.*

First, we need to split the metadata (containing the information of the sites and transects) and the community-matrix (containing the percent cover of each functional group on each transect) into two separate datasets before performing a PCoA.

```{r}
# create metadata
benthic.meta <- benthic.wide %>%
  select(Habitat, Transect)

# create percent cover dataset
benthic.base <- benthic.wide %>%
  ungroup() %>%
  select(-Habitat, -Transect)
```

Next, we need to create our distance matrix. This can be done using the `vegdist()` function which calculates pairwise dissimilarities between pairs of transects. Essentially, this will yield the absolute difference in percent cover across all functional groups — ecologically, it will represent the “total amount of compositional change between two transects.”

```{r}
# distance matrix using the Manhattan distance measure
benthic.dist <- vegdist(benthic.base, method = "manhattan")
```

The `wcmdscale()` function performs a Principal Coordinate Analysis. It takes the distance matrix and finds coordinates in a Euclidean space that best preserve those distances. `k = 2` → projects into 2 dimensions (for plotting). `eig = TRUE` → returns eigenvalues (variance explained per axis). Unlike Principal Component Analysis (PCA) that works directly on Euclidean distances, PCoA accepts any distance matrix (e.g., Manhattan, Bray-Curtis).

```{r}
benthic.pcoa.noscores <- wcmdscale(benthic.dist, k = 2, eig = TRUE)
summary(benthic.pcoa.noscores)
```

Now, we need to extract the scores, or the coordinates, for two different components of the ordination: 1) the species scores: scores for the functional groups (columns in the distance matrix) and 2) the site scores: scores for the transects (rows in the distance matrix).

**1: Species scores**

```{r}
benthic.pcoa <- BiodiversityR::add.spec.scores(benthic.pcoa.noscores, benthic.base, Rscale = TRUE, scaling = 1, multi = 1)

# view the projected species scores
benthic.pcoa$cproj
```

Then, we can compile this information into a tidy table by extracting the species loadings as a dataset. These loadings will be used to map the species onto the PCoA to visualize species-site relationships. 

```{r}
benthic.species.scores <- as.data.frame(benthic.pcoa$cproj) %>%
  mutate(species = rownames(.)) 
head(benthic.species.scores) 
```

**2: Site scores**

```{r}
benthic.site.scores <- as_tibble(benthic.pcoa$points, .name_repair = "minimal") %>% # these are the coordinates for each transect on the PCoA
  setNames(c("Dim1", "Dim2")) %>% 
  bind_cols(benthic.meta) # add in the metadata
head(benthic.site.scores)
```

One final thing we have to do before plotting is extract the information to plot the convex hulls of each site. Convex hulls enclose the points from each habitat into a polygon.

```{r}
benthic.hulls <- benthic.site.scores %>%
  group_by(Habitat) %>%
  dplyr::slice(chull(Dim1, Dim2))
head(benthic.hulls)
```

Now we have everything to plot our PCoA: species scores, site scores, and convex hulls.

```{r}
benthic.plot <- ggplot() +
  
   # plot species loadings as vectors
   geom_segment(data = benthic.species.scores,
    aes(x = 0, y = 0, xend = Dim1, yend = Dim2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "grey10",
    linewidth = 0.4) +
  
  # plot ellipses
  stat_ellipse(data = benthic.site.scores,
    aes(Dim1, Dim2, color = Habitat), type = "norm",
    linetype = 2) +
  
  # plot convex hulls
  geom_polygon(data = benthic.hulls,
    aes(Dim1, Dim2, fill = Habitat, group = Habitat),
    alpha = 0.6, color = "grey23",
    linewidth = 0.1) +
  
  # plot habitat coordinates
  geom_point(data = benthic.site.scores,
    aes(Dim1, Dim2, fill = Habitat, shape = Habitat),
    size = 3, alpha=0.9) +
  
  # add species labels
  ggrepel::geom_text_repel(
    data = benthic.species.scores,
    aes(Dim1, Dim2, label = species),
    fontface = "bold") +
  
  # aesthetics
  theme(
    axis.text = element_text(color = "black", size = 12),
    text = element_text(color = "black", size = 6)) +
  scale_color_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) + 
  scale_fill_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) +
  scale_shape_manual(values = c(21, 22, 23)) +
  labs(x = "Dimension 1", y = "Dimension 2") +
  theme_classic()

benthic.plot
```

**RESULTS:** The lack of overlap between convex hulls indicates that each reef habitat has a distinct benthic community. The species loadings show that the lagoon has a greater abundance of macroalgae, while the backreef has higher coral abundance.

## Generalized Linear Model

Now, we are interested in understanding the potential factors that drive coral abundance. For that, we will perform a Bayesian Generalized Linear Model using the package `{brms}`. Our model, in this case, will consist of macroalgae cover and habitat as predictor variables and coral cover as our response variable. We also will need to include transect as a random effect to account for quadrats taken within the same transect. Bayesian modeling was chosen over frequentist modeling to test the probability of the reef habitats being different from one another. 

To test whether we will need to use a general linear model or a generalized linear model, we must check the normality of coral abundance. 
```{r}
benthic.wide %>%
  ggplot(aes(x= Coral)) +
  geom_histogram() +
  theme_classic() +
  xlab("Coral Abundance") +
  ylab("Count")
```
From the histogram, you can see that coral abundance is **not** normally distributed, so we must model our data using a different distribution. The beta distribution is appropriate for this as our abundance data is continuous from 0-1.

Now we know that we need to run a generalized linear model using the argument `family =  beta.`
```{r}
# isolate model variables and convert character variables to factors
cm.data <- benthic.wide %>% 
  ungroup() %>%
  select(Habitat, Transect, Coral, Macroalgae) %>%
  mutate(Habitat = as.factor(Habitat),
         Transect = as.factor(Transect))
```

run the Bayesian model

```{r}
library(brms)
set.seed(690) # ensure that random sampling yields the same results every time

coral.glm <- brm(Coral ~ Habitat * Macroalgae + (1|Transect), data = cm.data, family = "beta")
summary(coral.glm)
```

Now lets explore the predictions of the model for the categorical variable

```{r}
set.seed(690)
habitat.effect <- cm.data %>%
  data_grid(Habitat = unique(Habitat),  # Use all unique levels of 'Site'
            Macroalgae = mean(Macroalgae, na.rm = TRUE)) %>%
  add_epred_draws(coral.glm, regex = T, ndraws = 200, re_formula = NA) %>%  # re_formula = NA for marginal effects
  ggplot(aes(x = Habitat, y = .epred, color = Habitat, fill = Habitat)) +
  stat_histinterval(alpha = 0.8, justification = -0.2) +
  theme_bw() +
  coord_cartesian(ylim = c(0, 0.3)) +
  xlab("Habitat") +
  ylab("Predicted coral cover (%)") +
  scale_color_manual(values = c("#FB9A29", "#CC6677", "#9970AB"))+
  scale_fill_manual(values = c("#FB9A29", "#CC6677", "#9970AB"))+
  theme(plot.margin = unit(c(0,0,0,0), 'lines')) +
  theme()

habitat.effect
```

Plot the effect of macroalgae cover for coral cover
```{r}
macroalgae.effect <- cm.data %>%
  data_grid(Macroalgae = seq(
  min(cm.data$Macroalgae, na.rm = TRUE),
  max(cm.data$Macroalgae, na.rm = TRUE),
  length.out = 100)) %>% 
  mutate(Habitat = "Backreef") %>% 
  add_epred_draws(coral.glm, regex = T, ndraws = 500, re_formula = NA)

   macroalgae.plot <- ggplot() +
  geom_line(data = macroalgae.effect, aes(x = Macroalgae, y = .epred, group = .draw), color = "#9970AB",   alpha = 1,linewidth = 0.1)+
  geom_point(data = cm.data, aes(y = Coral, x= Macroalgae), color = "#9970AB",
             fill = "#D3C0DD", alpha = 0.9,
            shape = 21, size = 3)+
  theme_bw() +
  theme_bw() +
  coord_cartesian(ylim = c(0, 0.2)) +
  xlab("Macroalgae cover (%)") +
  ylab("Predicted coral cover (%)") +
  theme(plot.margin = unit(c(0,0,0,0), 'lines')) +
  theme()

macroalgae.plot
```
