---
title: "Collaborative Data Science Project"
format: html
author: 
  - Juan Pablo Lozano Peña
  - Megan Do
  - Carly Dempsey
editor_options: 
  chunk_output_type: console
---

```{r}
#| output: false

knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)

# load in required packages
library(tidyverse)
```

## Research Question
##### *How does benthic cover vary across three reef zones of the Belize Barrier Reef?*

## Visualization of Data

#### 1) Metadata
This dataset contains the metadata for quadrat photos taken every 5m along 50m transects. Six transects were conducted at each of the three study sites in Belize: lagoon (whale shoal), fore reef, and back reef. The dataset contains the **image name**, **site of collection**, **date of collection**, **transect number**, and **quadrat number**.
```{r}
metadata <- read_csv("data/Belize_CoralNet_metadata.csv")
head(metadata)
```

#### 2) Percent Cover Dataset
This dataset consists of percent cover data for each quadrat photo. Percent cover was calculated using CoralNet, a software that automates the analysis of coral reef benthic images. Essentially, CoralNet generates 25 random points over each image, identifies/labels the points, and then calculates percent cover from that data. The dataset contains the **image ID**, **image name** (same as above), **annotation status** (whether the image has been labeled), **number of collection points**, and  various columns corresponding to **different benthic categories** such as hard coral species, soft coral species, macroalgal species, etc. See the [Glossary](#glossary) for a key to each category. 
```{r}
per.cover.data <- read_csv("data/Belize_CoralNet_percent.covers.csv")
head(per.cover.data)
```

## Data Wrangling

Combine the metadata with the percent cover data.
```{r}
# rename image name column to match that in the metadata
per.cover.data <- per.cover.data %>%
  rename("Image.name" = `Image name`)

# join the two together
benthic.data <- metadata %>%
  left_join(per.cover.data)
```

Now, we have to calculate the average percent cover for each category at each transect. To do this, we first must `gather()` our data so that we have a column for the categories and a column for the percent cover of each category.
```{r}
benthic.data <- benthic.data %>%
  gather(key = category, value = percent_cover, 9:69) # gather all category columns into one column, "category"
```

Next, we can group by Site, Transect, and Category to calculate the average percent cover for each category at each transect across the ten quadrats.
```{r}
transect.cover <- benthic.data %>%
  group_by(Site, Transect, category) %>%
  summarize(transect_cover = sum(percent_cover)/1000) # divide by 100 because there are 10 quadrats in a single transect: 100% * 10
head(transect.cover)
```
This dataset now shows the proportion of each category found in each transect. However, we also want to classify the categories into broad functional groups for downstream analyses.

```{r}
# classify functional groups
func.group.cover <- transect.cover %>%
  mutate(func_group = case_match(
    category,
    
    # macroalgae
    c("Brown_UFM", "Green_UFM", "Red_UFM", "Dictyota", "Halimeda", "Ulva", "Lobophora", "Sargassum", "Mac_matrix") ~ "Macroalgae",
    
    # corals
    c("AcrCerv", "AcrPalm", "AgAga", "AgTen", "AgUnd", "BL_OrbFav", "BORB", "CoNat", "DiplLab", "FaviaFrag", "MontaCav", "OrbAnn", "OrbFav", "OrbFrank", "PorAstr", "Por_branch", "PorPor", "PseuCliv", "PseuStrig", "Siderastr", "SidSid") ~ "Coral",
    
    # gorgonians
    c("Gorg_Fan", "Gorg_Plume", "Gorg_Whip", "GORG-BRI") ~ "Gorgonian",
    
    # turf
    c("EAM_HR", "EAM_LR", "EAM_RBL") ~ "Turf",
    
    # cyanobacteria
    c("Cyano_HR", "Cyano_LR", "Cyano_RBL", "Cyano_Sand") ~ "Cyanobacteria",
    
    # crustose coralline algae (CCA)
    c("CCA_LR", "CCA_HR", "CCA_RBL") ~ "CCA",
    
    # sponges
    c("Spong_Crus", "Spong_Erec", "Spong_Glo") ~ "Sponge",
    
    # seagrass
    c("SyrFil", "ThalTes") ~ "Seagrass",
    
    # encrusting algae
    c("Brown_Enc", "Red_Enc") ~ "Encrusting_algae",
    
    # abiotic
    c("Sand", "Bare_Rock", "Debris") ~ "Abiotic",
    
    # all other categories
    .default = "Other"))


# calculate percent cover for each functional group in each transect
func.group.cover <- func.group.cover %>%
  select(-category) %>%
  group_by(Site, Transect, func_group) %>%
  summarise(transect_cover = sum(transect_cover))
head(func.group.cover)
```


## Acknowledgements
This project draws on data generated through research conducted by Juan Pablo Lozano Peña.

## Glossary {#glossary}

* *Acropora cervicornis* (AcrCerv)
* *Acropora palmata* (AcrPalm)
* *Agaricia agaricites* (AgAga)
* *Agaricia tenuifolia* (AgTen)
* *Agaricia undata* (AgUnd)
* ##(BL_OrbFav)##
* ##(BORB)##
* *Colpophyllia natans* (CoNat)
* *Diploria labyrinthiformis* (DiplLab)
* *Favia fragum* (FaviaFrag)
* *Montastraea cavernosa* (MontaCav)
* *Orbicella annularis* (OrbAnn)
* *Orbicella faveolata* (OrbFav)
* *Orbicella franksi* (OrbFrank)
* *Porites astreoides* (PorAstr)
* ##(Por_branch)##
* *Porites porites* (PorPor)
* *Pseudodiploria clivosa* (PseuCliv)
* *Pseudodiploria strigosa* (PseuStrig)
* ##(Siderastr)##
* *Siderastrea siderea* (SidSid)
* sea anemones (ANEM)
* feather duster worms (Feat_dust)
* *Briareum* (GORG-BRI)
* sea fans (Gorg_Fan)
* sea plumes (Gorg_Plume)
* sea whips (Gorg_Whip)
* hydra (Hydra)
* fire corals (Millepora)
* ##(PCAR)##
* encrusting sponges (Sponge_Crus)
* ##(Sponge_Erec)##
* ##(Sponge_Glo)##
* ##(Film_Sand)##
* sand (Sand)
* bare rock (Bare_Rock)
* ##(CCA_LR)##
* ##(Cyano_HR)##
* ##(Cyano_LR)##
* ##(Cyano_RBL)##
* ##(Cyano_Sand)##
* debris (Debris)
* ##(ACA)##
* brown encrusting algae (Brown_Enc)
* brown algae ultra fine matrix (Brown_UFM)
* ##(CCA_HR)##
* ##(CCA_RBL)##
* *Dictyota* sp. (Dictyota)
* ##(EAM_HR)##
* ##(EAM_LR)##
* ##(EAM_RBL)##
* green algae ultra fine matrix (Green_UFM)
* *Halimeda* sp. (Halimeda)
* *Lobophora* sp. (Lobophora)
* macroalgae matrix (Mac_matrix)
* red encrusting algae (Red_Enc)
* red algae ultra fine matrix (Red_UFM)
* *Sargassum* sp. (Sargassum)
* *Ulva* sp. (Ulva)
* *Syringodium filiforme* (SyrFil)
* *Thalassia testudinum* (ThalTes)